HO unification from object language to meta language

### 2024-02-19 main.tex
thanks IA

\begin{abstract}
Specifying and implementing a logic from scratch requires a
considerable effort. Logical Frameworks and Higher Order Logic Programming
Languages provide dedicated, high level, Meta Languages (ML) to ease the
task in two ways: 1)
variable binding and substitution comes for free when ML binders
are used to represent object logic ones; 2)
proof construction, and even proof search, is greatly simplified
by leveraging the unification procedure provided by the ML.
Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and Isabelle~\cite{10.1007/978-3-540-71067-7_7}
that were used to implement various logics such as
First Order Logic~\cite{felty88cade},
Set Theory~\cite{10.1007/BF00881873} and
even Higher Order Logic~\cite{books/sp/NipkowPW02}.

The object logic we are interested in is Dependent Type Theory (DTT) for which
want to implement a unification procedure \U{} using
Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
Elpi comes equipped with the equational theory \Ue{} that comprises $\eta\beta$ equivalence
and higher order unification restricted to the pattern
fragment~\cite{miller92jsc}.
We want \U{} to feature the same equational theory of \Ue{}, but on the
object logic DTT. Unfortunately the natural encoding of DTT in Elpi, that we call \Fo{},
``underuses'' \Ue{} by making it applicable to first order unification problems only.
To overcome the problem we provide a better
behaved encoding \Ho{}, we show how to map a unification problems in \Fo{}
to related problems in \Ho{} and
how to map back the unifiers found by \Ue{}, effectively implementing
 \U{} on top of \Ue{} for the natural encoding \Fo{}.

We apply this technique to the implementation of a type-class~\cite{wadler89}
solver for Coq~\cite{Coq-refman}.
Type-class solvers are proof search procedures based on back-chaining designated
lemmas and provide essential automation to widely used 
Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
constitute our test bed.
\end{abstract}

>>>>

\begin{abstract}
Specifying and implementing a logic from scratch requires significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways: 1) variable binding and substitution are simplified when ML binders
represent object logic ones; 2) proof construction, and even proof search, is
greatly simplified by leveraging the unification procedure provided by the ML.
Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
which have been utilized to implement various formal systems such as
First Order Logic~\cite{felty88cade},
Set Theory~\cite{10.1007/BF00881873} and
even Higher Order Logic~\cite{books/sp/NipkowPW02}.

The object logic we are interested in is Dependent Type Theory (DTT),
for which we aim to implement a unification procedure \U{} using the ML
Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
Elpi comes equipped with the equational theory \Ue{}, comprising
$\eta\beta$ equivalence
and higher order unification restricted to the pattern
fragment~\cite{miller92jsc}.
We want \U{} to feature the same equational theory as \Ue{} but on the
object logic DTT.
Unfortunately the natural encoding of DTT in Elpi, which we refer to as \Fo{},
``underuses'' \Ue{} by restricting it to first-order unification problems only. 
To address this issue, we propose a better-behaved encoding, \Ho{},
demonstrate how to map unification problems in \Fo{}
to related problems in \Ho{}, and illustrate
how to map back the unifiers found by \Ue{}, effectively implementing
 \U{} on top of \Ue{} for the natural encoding \Fo{}.

We apply this technique to the implementation of a type-class~\cite{wadler89}
solver for Coq~\cite{Coq-refman}.
Type-class solvers are proof search procedures based on back-chaining designated
lemmas, providing essential automation to widely used 
Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
constitute our test bed.
\end{abstract}
